Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2023-02-14 paper/paper.tex
Inline synthesis

\section{Synthesis to CakeML}
\label{sec:synthesis}
\input{synthesis}

>>>>

%%
% SYNTHESIS TO CAKEML
%%

\section{Synthesis to CakeML}
\label{sec:synthesis}

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high assurance components specified by code contracts;
%
for each such occurrence it generates a CakeML program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

\paragraph{Regular expressions:\/} often the well-formedness of data on connections can be enforced by matching against a regular expression.
%
In other words, the data can be characterized by a \emph{regular language}.
%
As is well known, regular expressions can be translated to efficient deterministic finite state machine (DFA) implementations.
%
In \cite{formal-filter-synth-langsec,case-verified-filter} we report on the creation and application of table-based DFA filters from a verified regular expression-to-DFA compiler.
%
The latter paper offers a fully-worked out example synthesizing a regular expression filter and formally proving that it meets an non-terminating behavioral specification.

\paragraph{Contiguity types\cite{contiguity-types}:\/} contiguity types, similar to regular expressions, are effective at describing data at the network message level, \eg, as flat strings.
%
The notation is however much more expressive, being able to declare many kinds of so-called \emph{self-describing} messages, \ie, those where message structure is dependent on data held elsewhere in the message.
%
\figref{fig:filter-spec} gives a contiguity type specification for the well-formed waypoints of our example system.
%
We have formalized and verified a contiguity type matcher, which has been extensively applied throughout our case studies.

%%
% COMPILING FROM CODE CONTRACTS
%%

\subsection{Compiling from code contracts}

When full computational power is needed, we can use the expressive power of AGREE code contracts to write arbitrarily complex filter or monitor operations, as we have already seen.
%
Our formal model of AGREE syntax and semantics is applied to extract code contracts and map them to equivalent logic functions.
%
In particular, given a code contract with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics given in \secref{sec:code-contracts} can be applied to define a stream-oriented \emph{step} function $\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
%
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
%
Now suppose ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially imperative.
%
Then we can direcctly create a logic function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
%
corresponding to \konst{strmStep} that takes input and state as parameters and returns an updated state and output.
%
Thus the environment $\mathit{Env}$ holding all previous values of each variable can be replaced by \emph{parameters}, each of which is able to hold only one data element.
%
Assuming $\mathit{E}' = \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
%
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
%
In other words, projecting out the state and output components from the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of applying \konst{stateStep} on the input and state components of $\mathit{E}$ at time $t$.
%
This theorem connects the stream semantics of the code contract to a pure logic function over the values appearing in the stream.
%
From this point we can use tools in the \ckml\ eco-system to jump from the setting of higher order logic to \ckml\ abstract syntax \cite{cakeml-translator}.
%
In this way we have moved from stream semantics to logic function to CakeML evaluation semantics in a property-preserving way.
%
Once in \ckml, further necessary features can be added, namely imperative variables for storing state between component invocations, also invocations of the (verified) foreign function interface of \ckml\ \cite{cakeml-monadic}.
%
Finally, the CakeML compiler can be applied to the program.
%
It delivers an executable which can be shown to preserve the original contract.

%%
% GENERATING CAKEML
%%

\subsection{Generating \ckml}
For any of the three specification languages, we take the following general steps to arrive at an executable.
%
\begin{compactenum}
\item Map from specification to computable function in logic, producing a proof of equivalence.
\item Add parsers and printers (specified by contiguity types) for input and output.
\item Map to CakeML and add calls to input/output library routines via the CakeML foreign function interface.
\item Invoke the CakeML compiler.
\end{compactenum}
%
In the following, we examine further details in both filter and monitor synthesis.

%%
% FILTER GENERATION
%%

\subsection{Filter Generation}

A filter is intended to be simple, although it may make deep semantic checks.
%
A filter has one input port and one output; messages on the input that the filter policy admits pass unchanged to the output port;
%
all others are dropped (not passed on).
%
We have investigated two kinds of filter.
%
In the first, a relatively shallow scan of the input suffices to enforce the policy.
%
For example, we have used the expressive power of regular expressions and Contiguity Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints on GPS coordinates in UxAS messages.
%
On the other hand, a filter may need to parse the input buffer into a data structure specified in AGREE and apply a user-defined \emph{wellformedness} property, also specified in AGREE, to the data.
%
Arbitrarily complex wellformedness checks can be made in this way.
%
\figref{fig:filter-spec} shows a combination where the checking specified by {\small\verb+WELL_FORMED_AUTOMATION_RESPONSE+} depends on an underlying check specified by the contiguity type checking bounds on waypoints.

The verdict of a filter is made and performed within one thread invocation.
%
Thus, in its given time slice, the following steps must be completed:
\begin{compactenum}
\item
The filter checks to see if there is any input available.
%
If there is none then it yields control; otherwise:

\item The input is read (and parsed if need be);

\item The wellformedness predicate is evaluated on the input;

\item If the predicate returns \konst{true} then the input buffer
 is copied to the output, otherwise no action is taken; and

\item The filter yields control.
\end{compactenum}

\newsavebox{\contig}
\begin{lrbox}{\contig}
\begin{lstlisting}[style=myML]
  Waypoint =
    {Latitude  : f64
     Longitude : f64
     Altitude  : f32
     Check     : Assert
      (~90.0 <= Latitude and Latitude <= 90.0 and
       ~180.0 <= Longitude and Longitude <= 180.0 and
       1000.0 <= Altitude and Altitude <= 15000.0)}

  AutomationResponse =
    {TaskID : i64
     Length : u8
     Waypoints : Waypoint [3]}

 fun WELL_FORMED_AUTOMATION_RESPONSE(aresp) =
   (forall wpt in aresp.Waypoints, WELL_FORMED_WAYPOINT(wpt))
   and ... ;
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\contig}}
    \end{tabular}
  \end{center}
  \caption{Filter specification.}
  \label{fig:filter-spec}
\end{figure}

\newsavebox{\cml}
\begin{lrbox}{\cml}
\begin{lstlisting}[style=myML]
fun filter_step () =
 let val () = Utils.clear_buf buffer
     val () = API.callFFI "get_input" "" buffer
 in
    if WELL_FORMED_AUTOMATION_RESPONSE buffer
    then
      API.callFFI "put_output" buffer Utils.emptybuf
    else print"Filter rejects message.\n"
end
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\cml}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the filter.}
  \label{fig:filter-cakeml}
\end{figure}

The contiguity type specification and wellformedness predicate for the filter are shown in \figref{fig:filter-spec} and the synthesized \ckml\ code is in \figref{fig:filter-cakeml}.
%
The code is called at dispatch by the scheduler.
%
The \texttt{API.callFFI} is the link to the communication fabric to capture input and provide output.
%
The body of the function restates the filter contract to make the appropriate assignments in a way that matches the truth value of the predicate in the filter guarantee.

%%
% MONITOR GENERATION
%%

\subsection{Monitor Generation}

As discussed above, a monitor specification is mapped to a state transformation function
%
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}.
\]
%
which computes as follows:
%
\begin{compactenum}

\item Each input buffer is parsed into data of the type specified by the port type;

\item
New values for the state variables are computed, in dependency order.
%
At this point variables defined by \emph{followed-by} are dealt with.
%
Suppose the variable definitions have the following form:
%
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
%
In the generated function, for the first invocation of \konst{stateStep} only, the initializations are executed in order:
%
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
%
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
%
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]

\item Values of the outputs are computed;

\item Outputs are written and the new state is written;

\item The monitor yields control.
\end{compactenum}

The \konst{stepFn} for the monitor of the example described in \secref{sec:example} is displayed in \figref{fig:monitor-cakeml}.

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
fun stateStep (response,request)
    (rsp,req,preIsPending,isPending,latency,policy,alert)
let val stateVars' =
      if !initStep then
      let val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = False
          val isPending = req andalso not rsp
          val latency = 0
          val policy = (if rsp then req else True)
          val alert = not(policy)
          val () = (initStep := False)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
      else
      let ...
          val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = pre isPending
          val isPending = (req andalso not rsp) orelse
                          ((not rsp) andalso pre isPending)
          val latency = (if req then 0 else (pre latency + 1)
          val policy =
            ((if isPending then
                (latency < Defs.max_latency) else True)
              andalso
              (if rsp then
                 (req orelse preIsPending) else True))
          val alert = (Defs.is_latched andalso (pre alert))
                      orelse not(policy)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
   val (rsp',_,_,_,_,_,alert') = stateVars'
   val alertPort = if alert' then Some () else None
   val outputPort =
     if alert' then None else
     if rsp' then Some response
     else None
in
   (stateVars', (alertPort,outputPort))
end
\end{lstlisting}
\end{lrbox}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the monitor.}
  \label{fig:monitor-cakeml}
\end{figure}

\subsection{Incorporation into system execution}

The system scheduler \emph{activates} components in some order.
%
It is an obligation on the system that the scheduler follows some sensible partial order of component activation and allows each component sufficient time for its computation.
%
Activating a monitor component takes the form of the following pseudo-code, in which the monitor evaluates the \konst{stateStep} on its current inputs and the current values of the state variables, returning the new state and the output values.
%
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stateStep} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\paragraph{Partiality:\/} partiality is an important consideration for both filters and monitors: the input data might not parse or the computation of a filter could be badly written and fail at runtime.
%
In such cases, the filter should recover and yield control without passing the input onwards.
%
For these failure cases, the filter is behaving as it should, but we must also guard against situations in which a \emph{correctly specified} filter or monitor fails in the runtime.
%
This kind of defect arises when a filter, for example, \emph{ought} to accept a message, but lack of resources results in the filter failing to do so.
%
For example, the parse of a message might need more space than has been allocated;
%
another example could be if the time slice provided by the scheduler is too short for the computation to finish.
%
Thus resource bounds need to be included in the correctness argument.
%
Some preliminary work that could support formal proofs of such bounds has been done for \ckml\ \cite{cakeml-space-cost}.
